GSODTools
=========

### Global options

To be honest, I have no idea how to hide the following lines of code that take 
care of setting some global options. The file README.rmd is generated using the
`knitr` package, and it works out fine calling 'Knit HTML' from within RStudio, 
i.e. code chunks defined as `include = FALSE` are not displayed. However, when I 
upload the file to GitHub, the package's front page shows all code chunks in 
README.rmd, both `include = TRUE` and `include = FALSE`. Any help on that would 
be highly appreciated. Anyway, just ignore the following except 
you are interested in setting some global options.

```{r global_options, echo = FALSE, include = FALSE}
library(knitr)
options(width = 120)
opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'Figs/',
               include = TRUE, warning = FALSE, message = FALSE)
```

### What it is all about

But to return to the actual topic: Every person dealing with long-term 
climatological data (e.g. of daily air temperature, relative humidity, and precipitation amounts) will sooner or later stumble across the Global Summary Of Day (GSOD) climate data collection provided by the National Oceanic and Atmospheric Association (NOAA). I've been 
recently looking for available GSOD stations in close vicinity to Mt. Kilimanjaro, Tanzania, and as I am trying to realize most of my coding work using R, I quickly 
noticed that there are only a few packages that provide convenient tools for 
processing GSOD data. Therefore, I started to write this package that includes both downloading data sets of selected climate stations for a given time span as well as 
some processing steps for quality assurance and gap filling.


### Introducing the processing chain

**Getting started**

The starting point for each GSOD-related search query is the selection of a 
particular station (or even multiple stations). Although a [GIS Data Locator][1] exists that allows interactive 
station selection and data acquisition, I thought it was a good thing to 
implement a couple of search function to speed things up a little bit. 

The **GSODTools** package comes with a built-in dataset from [NOAA's FTP server][2]
holding information about all available GSOD stations that is automatically 
attached via lazy-loading when loading the package. Let's have a quick look at it. 

```{r gsodstations, echo = FALSE}
library(GSODTools)
head(gsodstations)
```

Unfortunatelly, the data formatting and consistency of this official table is 
rather poor. Hence, I thought it might be quite helpful to sort out some 
inconveniences above all else. The referring function is called *gsodReformat()* 
and allows to reformat elevation (decimeters to meters) as well as latitude and longitude
(thousandth of a degree to whole degree). Furthermore, it offers the opportunity
to remove invalid coordinates, i.e. values of latitude and longitude exceeding 
valid thresholds of +/-90 and +/-180 degrees, respectively. Optionally, the 
adjusted dataset can be converted to an object of class `sp` prior to return.

Consequently, the first lines of code working with **GSODTools** should probably 
look like this.

```{r gsodReformat}
# Reformat data and convert to spatial object
gsod_shp <- gsodReformat(data = gsodstations,
                         elevation = TRUE, 
                         coords = TRUE,
                         df2sp = TRUE)

par(mar = c(0, 0, 0, 0))
plot(gsod_shp)
```

**Downloading data**

Now that the list of available GSOD stations is in a reasonable format and holds
spatial information, the next step would be to select a station you would like to
download data from. Using the GIS Data Locator, this involves quite some clicking 
around until you finally reach the download page. **GSODTools** offers multiple 
functions to facilitate station selection and data acquisition, allowing the user
to select stations based on spatial characteristics or by hand. 

`stationFromCoords` takes a x (longitude) and y (latitude) coordinate as input, 
and returns all available GSOD stations that fall within a user-defined buffer
around that location. Alternatively, a 'SpatialPoints' object may be provided 
rather than two separate numerics. For instance, let's search for GSOD stations
in a circle of 500 km around Kibo summit, Mt. Kilimanjaro, Tanzania. The referring
coordinates are `c(37.359031, -3.065053)`.

```{r stationFromCoords}
shp_kibo <- stationFromCoords(x = 37.359031, y = -3.065053, width = 500)
# or: stationFromCoords(x = c(37.359031, -3.065053), width = 500)
# or: stationFromCoords(x = SpatialPoints(data.frame(x = 37.359031, 
#                                                    y = -3.065053), 
#                                         proj4string = CRS("+init=epsg:4326")), 
#                       width = 500)

mapGriddedData(mapRegion = "africa", plotData = FALSE, borderCol = "black",
               addLegend = FALSE)
points(shp_kibo, col = "red", pch = 20, cex = 2)
```

`stationFromExtent`, just like `stationFromCoords`, allows station selection 
based on spatial criteria. However, the user is prompted to manually draw an 
extent on a map rather than directly supplying specific coordinates. The advantage 
is that spatial selection is not performed in a circular shape, i.e. in a uniform
distance around a given location, but depends on user preferences. With respect
to the aforementioned example, this means that GSOD stations in the southern
Mt. Kilimanjaro region could be selected rather than all stations in a given 
distance from the summit. Alternatively, a rectangular bounding box can be 
supplied instead of calling `drawExtent` (which is actually quite difficult to
include in a README file). 

```{r stationFromExtent}
bbox_kibo_south <- extent(c(36.6, 37.72, -3.5, -3.065053))
shp_kibo_south <- stationFromExtent(bb = bbox_kibo_south)

mapGriddedData(mapRegion = "africa", plotData = FALSE, borderCol = "black",
               addLegend = FALSE)
points(shp_kibo_south, col = "red", pch = 20, cex = 2)
```


[1]: http://www.climate.gov/daily-observational-data-global-summary-day-gsod-%E2%80%93-gis-data-locator
[2]: ftp://ftp.ncdc.noaa.gov/pub/data/gsod/ish-history.csv
