GSODTools
=========

```{r global_options, include = FALSE}
library(knitr)
options(width = 120)
opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'Figs/',
               include = TRUE, warning = FALSE, message = FALSE)
```

### What it is all about

Every person dealing with long-term climatological data (e.g. of daily air
temperature, relative humidity, and precipitation amounts) will sooner or later
stumble across the Global Summary Of Day (GSOD) climate data collection 
provided by the National Oceanic and Atmospheric Association (NOAA). I've been 
recently looking for available GSOD stations in close vicinity to Mt. Kilimanjaro, Tanzania, and as I am trying to realize most of my coding work using R, I quickly 
noticed that there are only a few packages that provide convenient tools for 
processing GSOD data. Therefore, I started to write this package that includes both downloading data sets of selected climate stations for a given time span as well as 
some processing steps for quality assurance and gap filling.


### Introducing the processing chain

**Getting started**

The starting point for each GSOD-related search query is the selection of a 
particular station (or even multiple stations). Although a [GIS Data Locator][1] exists that allows interactive 
station selection and data acquisition, I thought it was a good thing to 
implement a couple of search function to speed things up a little bit. 

The **GSODTools** package comes with a built-in dataset from [NOAA's FTP server][2]
holding information about all available GSOD stations that is automatically 
attached via lazy-loading when loading the package. Let's have a quick look at it. 

```{r gsodstations, echo = FALSE}
library(GSODTools)
head(gsodstations)
```

Unfortunatelly, the data formatting and consistency of this official table is 
rather poor. Hence, I thought it might be quite helpful to sort out some 
inconveniences above all else. The referring function is called *gsodReformat()* 
and allows to reformat elevation (decimeters to meters) as well as latitude and longitude
(thousandth of a degree to whole degree). Furthermore, it offers the opportunity
to remove invalid coordinates, i.e. values of latitude and longitude exceeding 
valid thresholds of +/-90 and +/-180 degrees, respectively. Optionally, the 
adjusted dataset can be converted to an object of class `sp` prior to return.

Consequently, the first lines of code working with **GSODTools** should probably 
look like the following.

```{r gsodReformat}
# Reformat data and convert to spatial object
gsod_shp <- gsodReformat(data = gsodstations,
                         elevation = TRUE, 
                         coords = TRUE,
                         df2sp = TRUE)

par(mar = c(0, 0, 0, 0))
plot(gsod_shp)
```

**Downloading data**


[1]: http://www.climate.gov/daily-observational-data-global-summary-day-gsod-%E2%80%93-gis-data-locator
[2]: ftp://ftp.ncdc.noaa.gov/pub/data/gsod/ish-history.csv