---
title: "GSODTools"
output: md_document
---

```{r global_options, include = FALSE}
library(knitr)
options(width = 120)
opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'figure/',
               include = TRUE, warning = FALSE, message = FALSE)
```

For the latest stable release, please refer to  
[![doi_GSODTools_v1.0.0][1]][2].

### What it is all about

But to return to the actual topic: Every person dealing with long-term 
climatological data (e.g. of daily air temperature, relative humidity, and precipitation amounts) will sooner or later stumble across the Global Summary Of Day (GSOD) climate data collection provided by the National Oceanic and Atmospheric Association (NOAA). I've been 
recently looking for available GSOD stations in close vicinity to Mt. Kilimanjaro, Tanzania, and as I am trying to realize most of my coding work using R, I quickly 
noticed that there are only a few packages that provide convenient tools for 
processing GSOD data. Therefore, I started to write this package that includes both downloading data sets of selected climate stations for a given time span as well as 
some processing steps for quality assurance and gap filling.


### Introducing the processing chain

**Getting started**

The starting point for each GSOD-related search query is the selection of a 
particular station (or even multiple stations). Although there are tools that 
allow interactive station selection and data acquisition, I thought it was a 
good thing to implement a couple of search functions to speed things up a little
bit. 

The **GSODTools** package comes with a built-in dataset from [NOAA's FTP server][3]
holding information about all available GSOD stations that is automatically 
attached via lazy-loading when loading the package. Let's have a quick look at it. 

```{r gsodstations, echo = FALSE}
library(GSODTools)
head(gsodstations)
```

To transform the built-in dataset into a spatial object, either run 
`sf::st_as_sf()` manually or use the convenience function `gsodDf2Sp()`. Past 
inconveniences (elevation in decimeters, coordinates in thousandth of a degree, 
longitude and latitude outside of -180 to 180 and -90 to 90, respectively) are 
no longer an issue, which is why `gsodReformat()` has been deprecated and will 
eventually be removed from the package.

```{r gsodDf2Sp}
# Reformat data and convert to spatial object
gsod_shp <- gsodDf2Sp(data = gsodstations)

par(mar = c(0, 0, 0, 0))
plot(gsod_shp)
```

**Selecting a station**

Now that the list of available GSOD stations is in a reasonable format and holds
spatial information, the next step would be to select a station you would like to
download data from. Using the GIS Data Locator, this involves quite some clicking 
around until you finally reach the download page. **GSODTools** offers multiple 
functions to facilitate station selection and data acquisition, allowing the user
to select stations based on spatial characteristics or by hand. 

`stationFromCoords` takes a x (longitude) and y (latitude) coordinate as input, 
and returns all available GSOD stations that fall within a user-defined buffer
around that location. Alternatively, a 'SpatialPoints' object may be provided 
rather than two separate numerics. For instance, let's search for GSOD stations
in a circle of 500 km around Kibo summit, Mt. Kilimanjaro, Tanzania. The referring
coordinates are `c(37.359031, -3.065053)`.

```{r stationFromCoords}
library(mapview)

shp_kibo <- stationFromCoords(x = 37.359031, y = -3.065053, width = 500)
# or: stationFromCoords(x = c(37.359031, -3.065053), width = 500)
# or: stationFromCoords(x = SpatialPoints(data.frame(x = 37.359031, 
#                                                    y = -3.065053), 
#                                         proj4string = CRS("+init=epsg:4326")), 
#                       width = 500)

mapview(
  shp_kibo
)

```

`stationFromExtent`, just like `stationFromCoords`, allows station selection 
based on spatial criteria. There are actually two options how to handle this 
function. If no bounding box is defined, the user is automatically prompted to 
manually draw an extent on a map rather than directly supplying specific 
coordinates. The advantage is that spatial selection is not performed in a 
circular shape, i.e. in a uniform distance around a given location, but depends 
on user preferences. With respect to the aforementioned example, this means that GSOD stations in the southern Mt. Kilimanjaro region could be selected rather than all stations in a given distance from the summit. 
Alternatively, an extent object from an arbitrary spatial object, e.g. 'RasterLayer', 
'SpatialPolygons' etc, may be defined. In this case, `drawExtent` (which is actually quite difficult to include in a README file) is automatically disabled. 

```{r stationFromExtent}
bbox_kibo_south <- sf::st_bbox(
  c(
    xmin = 36.6
    , xmax = 37.72
    , ymin = -3.5
    , ymax = -3.065053
  )
)
shp_kili_south <- stationFromExtent(bb = bbox_kibo_south)

mapview(
  shp_kili_south
)

```

The third and, at the moment, final possibility to select a GSOD station is to 
simply choose a name from the built-in station list. This is, however, a quite
tricky approach since you have to know the precise spelling of a station's name. 
Again referring to the above example where we selected Arusha, Moshi, and 
Kilimanjaro International Airport (KIA), this would more or less look like this. 

```{r station_manual_selection}
station_names <- c("ARUSHA", "KILIMANJARO INTL", "MOSHI")

(
  shp_kili_south <- 
    gsodstations |> 
    gsodReformat() |> 
    subset(`STATION NAME` %in% station_names) |> 
    gsodDf2Sp()
)

```

**Downloading data**

Ideally, you have now found an appropriate station you would like to acquire 
data from. It usually takes some patience to click through the download procedure
inherent to the GIS Data Locator, and having to perform the same steps again and 
again can soon get quite frustrating. I implemented a function called 
`dlGsodStations` that helps to overcome these inconveniences. 
The function works with USAF codes that are unique to each GSOD station and, in 
my opinion, catchier than the associated station names. The USAF code can be 
determined by having a look at the outcome of the various station selection 
functions. 

```{r usaf_retrieval}
# Subset station list by name, and display related USAF code
moshi <- subset(
  gsodstations
  , `STATION NAME` == "MOSHI"
)
head(moshi)
```

If you are not willing to download the entire dataset from a given 
station (which is the default setting), but rather a limited period of time, you 
can specify the desired years through `start_year` and `end_year`. It is 
possible to adjust the destination folder for data download (`dsn`) if you do 
not want to save the data in the current working directory. Furthermore, you can
also determine whether to extract the zipped files and remove the *.gz files
after extraction. In this case, the function returns a `data.frame` rather than
a vector of *.gz filepaths. Note that, for each year and station, the function 
automatically checks if data is already present in the specified destination 
folder. If so, the referring download will be skipped and the 
subsequent download query will be evaluated. Here is an example on data 
download including visualization of the daily mean air temperature from Moshi, 
Tanzania, for the years 1990 to 2000. 

```{r dlGsodStations_echo, results = "hide", eval = FALSE}
# Not run: Download data from Moshi, Tanzania, from 1990 to 1995
gsod_moshi <- dlGsodStations(usaf = moshi$USAF,
                             start_year = 1990, end_year = 1997,
                             dsn = tempdir(),
                             unzip = TRUE)
```

```{r dlGsodStations_eval, results = "hide", echo = FALSE}
# Not run: Download data from Moshi, Tanzania, from 1990 to 1995
gsod_moshi = dlGsodStations(
  usaf = moshi$USAF
  , start_year = 1990
  , end_year = 1997
  , dsn = file.path(
    "inst/extdata"
    , moshi$USAF
  )
  , unzip = TRUE
)

```

```{r dlGsodStations_visualize}
# Plot temperature data (but: time series not continuous!)
library(ggplot2)

# Convert temperature column
gsod_moshi$TEMP <- toCelsius(gsod_moshi$TEMP, digits = 1)

ggplot(aes(y = TEMP, x = YEARMODA), data = gsod_moshi) + 
  geom_line(color = "grey50") + 
  stat_smooth(method = "lm") + 
  labs(x = "Index", y = expression("Temperature (" ~ degree ~ C ~ ")")) +
  theme_bw()
```

**Side note: `toCelsius`**

You may have already noticed `toCelsius` in the preceding code chunk. Indeed, 
this function is as small as useful, as it converts temperature values from 
degree Fahrenheit (which is native GSOD format) to degree Celsius. 

```{r toCelsius}
# RNG
set.seed(10)

# Degree Fahrenheit
val_fah <- runif(10, 60, 80)
# -> Degree Celsius
toCelsius(val_fah, digits = 1)
```

[1]: https://zenodo.org/badge/5994/environmentalinformatics-marburg/GSODTools.png
[2]: http://dx.doi.org/10.5281/zenodo.12217
[3]: ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv
